### 2015-01-15 doc/thesis/system.tex
Fixing the text about tables subtyping

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of the tables that have no keys that do not
inhabit one of the table's key types, and that have no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and other table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{u} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open|unique}}
\end{array}
\]

Unlike the rule \textsc{S-TABLE1}, the rule \textsc{S-TABLE2} allows
covariant keys and do not allow width subtyping.
We allow this behavior to \emph{unique} table types because we also
want to use them as a way to join table fields.
For instance, we may want to use the table constructor to initialize
a variable with a table type that describes a hash.

@@

We also do not allow width and depth subtyping between \emph{open}
table types or between \emph{open} and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\forall i \in 1..m \; \senv \vdash k_{i} \subtype k_{i}' \;\;\;
       \senv \vdash v_{i} \subtype_{c} v_{i}' \;\;\;
       \forall j \in m+1..n \; \senv \vdash \Nil \subtype_{o} v_{j}'}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{m}'{:}v_{m}', ..., k_{n}'{:}v_{n}'\}_{closed|open}}
\end{array}
\]

The rule \textsc{S-TABLE3} uses the same rules from
\textsc{S-TABLE1} for handling the type of the values of each key, and
it uses the same rules from \textsc{S-TABLE2} to allow joining fields plus
omitting optional fields.

@@

Besides allowing the refinement of table types through field assignment,
Typed Lua also keeps track whether it is safe or not safe to change the
type of a table.
For instance,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}
does not type check because aliasing $a$ produces the type
$\{``foo":\Integer\}_{closed}$ that is not subtype of
$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
The rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2} define this behavior:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-IDREAD1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{unique}}
      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}[n \mapsto \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}]}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IDREAD2}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{open}}
      {\env_{1} \vdash n_{e}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}_{closed}, \env_{1}}
\end{array}
\end{array}
\]

Aliasing \emph{unique} and \emph{open} table types can be unsafe.
For this reason, the rules \textsc{T-IDREAD1} and \textsc{T-IDREAD2}
promote the type of an alias expression from \emph{unique} or \emph{open}
to \emph{closed}.
The rule \textsc{T-IDREAD1} also promotes the aliased variable from
\emph{unique} to \emph{open}, as \emph{unique} table types only
denote the type of table types that have no alias.

@@

We also have different rules for type checking table indexing that
appear as expressions to avoid promoting an \emph{unique} table to
\emph{open} in these operations:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEXREAD1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
       \env_{1} \vdash e_{2}:t, \env_{2} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash n_{e}[e_{2}]:v_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEXREAD1}\\
\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\end{array}
\]

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{closed} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t}
      {\env_{1} \vdash {<}t{>} \; n:close(t), \env_{1}[n \mapsto t]}
\end{array}
\]

>>>>

We introduced \emph{unique} table types to avoid this limitation,
as they represent the type of the tables that have no keys that do not
inhabit one of the table's key types, and that have no alias.
In particular, this is the case of the table constructor.
The following subtyping rule defines the subtyping relation among
\emph{unique} table types and \emph{closed} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE2}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\]

The rule \textsc{S-TABLE2} allows width subtyping and covariant keys.
It allows covariant keys because we also want to use \emph{unique}
table types as a way to join table fields that inhabit \emph{closed} table types.
For instance, we want to use the table constructor to initialize
a variable with a table type that describes a hash.

@@

Even though we allow width subtyping between \emph{unique} and \emph{closed}
table types, we do not allow it among \emph{unique} and \emph{open} or
\emph{unique} table types because it would violate our definition of
\emph{unique} and \emph{open} table types:
\[
\begin{array}{c}
\mylabel{S-TABLE3}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{u} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{unique} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open|unique}}
\end{array}
\]

The rules that handle subtyping among \emph{open} table types and
\emph{open} or \emph{closed} table types are similar to the rules
that handle subtyping among \emph{unique} table types and other table types:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-TABLE4}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \; \forall j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{closed}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-TABLE5}\\
\dfrac{\begin{array}{c}
       \forall i \in 1..m \\
       \exists j \in 1..n \;
       \senv \vdash k_{i} \subtype k_{j}' \land \senv \vdash v_{i} \subtype_{c} v_{j}' \\
       \forall j \in 1..n \; \not\exists i \in 1..m \;
       \senv \vdash k_{i} \subtype k_{j}' \to \senv \vdash \Nil \subtype_{o} v_{j}'
       \end{array}}
      {\senv \vdash \{k_{1}{:}v_{1}, ..., k_{m}{:}v_{m}\}_{open} \subtype
                    \{k_{1}'{:}v_{1}', ..., k_{n}'{:}v_{n}'\}_{open}}
\end{array}
\end{array}
\]

The rule \textsc{S-TABLE4} allows width subtyping, while
the rule \textsc{S-TABLE5} does not allow width subtyping.
Both rules allow joining fields plus omitting optional fields.
Both rules use $\subtype_{c}$ to allow depth on $\Const$ fields only.

@@

Besides allowing the refinement of table types through field assignment,
Typed Lua also keeps track whether it is safe or not safe to change the
type of a table.
For instance,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{``foo":\Integer\}_{unique} = \{ [``foo"] = 5 \} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{``foo":\Integer \cup \Nil \}_{closed} = a \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b[``foo"] = \mathbf{nil}$}
\end{tabular}
\end{center}
does not type check because aliasing $a$ produces the type
$\{``foo":\Integer\}_{closed}$ that is not subtype of
$\{``foo":\Integer \cup \Nil\}_{closed}$, the type of $b$.
The rule \textsc{T-IDREAD} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-IDREAD}\\
\dfrac{\env_{1}(n) = t}
      {\env_{1} \vdash n_{e}:close(t), \env_{1}[n \mapsto open(t)]}
\end{array}
\]

Aliasing \emph{unique} and \emph{open} table types can be unsafe.
For this reason, the rule \textsc{T-IDREAD} promote the type of an
alias expression from \emph{unique} or \emph{open} to \emph{closed},
and it also promotes the aliased variable from
\emph{unique} to \emph{open}, as \emph{unique} table types only
denote the type of table types that have no alias.

@@

We also need to close \emph{unique} and \emph{open} tables that
appear in the left-hand side of assignments.
For instance,
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{open} = \{\} \; \mathbf{in}$}\\
& & \multicolumn{1}{l}{$b = a;$}\\
& & \multicolumn{1}{l}{$a[``x"] \; {<}\String{>} = ``foo";$}\\
& & \multicolumn{1}{l}{$b[``x"] \; {<}\Integer{>} = 1$}\\
\end{tabular}
\end{center}
does not type check because we cannot add the field $``x"$ to $b$,
as its type is \emph{closed}, and thus does not allow changing the
value that is stored in the field $``x"$ of local $a$.
The rule \textsc{T-IDWRITE} defines this behavior:
\[
\begin{array}{c}
\mylabel{T-IDWRITE}\\
\dfrac{\env_{1}(n) = t}
      {\env_{1} \vdash n_{l}:close(t), \env_{1}[n \mapsto close(t)]}
\end{array}
\]

We also have different rules for type checking table indexing to avoid
promoting table types in these operations:
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-INDEX1}\\
\dfrac{\env_{1}(n) = \{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\} \;\;\;
       \env_{1} \vdash e_{2}:t, \env_{2} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash n[e_{2}]:v_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-INDEX2}\\
\dfrac{\env_{1} \vdash e_{1}:\{k_{1}{:}v_{1}, ..., k_{n}{:}v_{n}\}, \env_{2} \;\;\;
       \env_{2} \vdash e_{2}:t, \env_{3} \;\;\;
       \exists i \in 1{..}n \; t \lesssim k_{i}}
      {\env_{1} \vdash e_{1}[e_{2}]:v_{i}, \env_{3}}
\end{array}
\end{array}
\]

Sometimes we want to use a local variable that holds an \emph{unique}
table type to initialize a \emph{closed} table type, as \emph{unique}
table types allow us to initialize optional fields.
To do that, Typed Lua includes a cast expression that allow us
to use an \emph{unique} table type before it becomes \emph{open}.
The rule \textsc{T-CAST} describes this behavior:
\[
\begin{array}{c}
\mylabel{T-CAST}\\
\dfrac{\env_{1}(n) \subtype t \;\;\;
       \env_{1}[n \mapsto t] \vdash n_{e}:t_{1}, \env_{2}}
      {\env_{1} \vdash {<}t{>} \; n:t_{1}, \env_{2}}
\end{array}
\]
